---
title: "Mapping the Pandemic Landscape"
author: "Evan Kennedy elk2149, Amrutha Banda ab6169, Kaveri Uberoy kiu2103, Suhina Sharma ss7636"
output: html_document
---

```{r echo= FALSE}
knitr::opts_chunk$set(
echo = FALSE,
message = FALSE,
warning = FALSE
)
```

![](https://www.whitehousetx.org/ImageRepository/Document?documentId=480)


### Screencast

<!-- YouTube Embed -->
<div style="position:relative;padding-bottom:56.25%;height:0;overflow:hidden;max-width:100%;margin-top:20px;margin-bottom:20px;">
  <iframe 
    src="https://www.youtube.com/embed/TXgjqQ-aa_s" 
    title="YouTube video player"
    frameborder="0"
    allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share"
    allowfullscreen
    style="position:absolute;top:0;left:0;width:100%;height:100%;">
  </iframe>
</div>



### Motivation:

The Covid-19 pandemic unfolded unevenly across the United States, shaped not only by viral transmission but also by social, political, and structural forces. In this project, we examine county-level Covid-19 cases and deaths alongside factors such as race, poverty, political affiliation, and mask use to better understand how these contextual variables influenced pandemic outcomes. We also explore the relationship between Covid-19 incidence and violence, considering how social stressors and community-level vulnerability may have interacted with the spread of the virus.

### Intended Project:

This project provides a county-level analysis of how COVID-19 cases and deaths varied across the U.S. in relation to social and structural factors—including race, poverty, political affiliation, mask use, and community violence. Our final products include statistical summaries and an interactive Shiny map that allows users to explore yearly COVID trends alongside these demographic and socioeconomic variables. Together, these tools highlight how COVID-19 outcomes intersected with broader social and political conditions across the country.

### Our anticipated data sources:
Datasets: [socioeconomic data](https://github.com/evangambit/JsonOfCounties), COVID-19 data ([2020](https://github.com/nytimes/covid-19-data/blob/master/us-counties-2020.csv),[2021](https://github.com/nytimes/covid-19-data/blob/master/us-counties-2021.csv),[2022](https://github.com/nytimes/covid-19-data/blob/master/us-counties-2022.csv),[2023](https://github.com/nytimes/covid-19-data/blob/master/us-counties-2023.csv), [recent](https://github.com/nytimes/covid-19-data/blob/master/us-counties-recent.csv)). These datasets draw from reputable government and news and research institutions, as described on each page. Together, joined by county, these sources will enable a detailed, multidimensional view of the socioeconomic and environmental factors correlated with the COVID-19 outcomes across U.S. counties.

### Planned Analysis:
1. Clean, merge, and standardize county-level datasets; calculate cumulative COVID-19 incidence and mortality rates; examine temporal trends across U.S. regions.
2. Model relationships between COVID-19 outcomes and county characteristics, including environmental factors and demographic variables.

### Planned Visuals:	
An interactive U.S. map displaying case and death rates alongside socioeconomic and environmental factors, time-series plots of cases by region, and scatterplots illustrating relationships between COVID-19 outcomes and key county-level factors.

### Anticipated Challenges
Harmonizing datasets, creating mapped visuals, dealing with co-linearity between factors, managing time lag in correlation with covid incidence.

```{r}
library(htmltools)

HTML('
<div style="background:#f7f7f7; padding:15px; border-radius:6px; margin-top:25px;">

<h3>Interactive Case/Death Map</h3>
<p>See an interactive map of covid cases and deaths across the county:</p>
<p><strong><a href=https://ekennedycu.shinyapps.io/shiny_map/>View→</a></strong></p>

</div>
')
```

```{r}
HTML('
<div style="background:#f7f7f7; padding:15px; border-radius:6px; margin-top:25px;">

  <h3>Covid-19 and Violence</h3>
  <p>See an investigation of Covid-19 and the incidence of community violence:</p>
  <p><strong><a href="violence_covid.html">View Analysis →</a></strong></p>

</div>
')
```

```{r}
HTML('
<div style="background:#f7f7f7; padding:15px; border-radius:6px; margin-top:25px;">

  <h3>Covid, Poverty, and Race</h3>
  <p>See an investigation of Covid-19, poverty, and race:</p>
  <p><strong><a href="Poverty_Race_COVID_final.html">View Analysis →</a></strong></p>

</div>
')
```

```{r github_buttons, echo=FALSE, results='asis'}
HTML('
<div style="background:#f7f7f7; padding:15px; border-radius:6px; margin-top:25px; text-align:center;">

  <h3>GitHub Repositories</h3>

  <p style="margin-top:10px;">
    <strong><a href="https://github.com/kaveriuberoy/p8105_final_project" target="_blank">Project Repo →</a></strong>
  </p>

  <p>
    <strong><a href="https://github.com/ekennedyCU/p8105_finalproject_shinymap" target="_blank">Shiny Map Repo →</a></strong>
  </p>

</div>
')
```

