---
title: "Poverty & Race and COVID"
output: 
  html_document:
    theme: cosmo
date: "2025-11-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

```{r}
covid_demo_final_df <- readRDS("covid_demo_final_df.rds")
covid_demo_final_df
```


```{r}
final_df_1 = covid_demo_final_df |> 
  select(county, state, race_non_hispanic_white_alone_male, race_non_hispanic_white_alone_female, race_black_alone_male, race_black_alone_female, race_asian_alone_male, race_asian_alone_female, race_hispanic_male, race_hispanic_male, race_hispanic_female, health_percent_children_in_poverty,avg_income, poverty_rate, health_80th_percentile_income, health_20th_percentile_income, year, yearly_cases, yearly_deaths, population_2019)

final_df_1
```
```{r}
final_df <- final_df_1 |> 
  mutate(year = as.character(year)) |> 
  pivot_wider(
    id_cols = c(county, state, race_non_hispanic_white_alone_male, race_non_hispanic_white_alone_female, 
                race_black_alone_male, race_black_alone_female, race_asian_alone_male, race_asian_alone_female, 
                race_hispanic_male, race_hispanic_female, health_percent_children_in_poverty, avg_income, 
                poverty_rate, health_80th_percentile_income, health_20th_percentile_income, population_2019),
    names_from = year,
    values_from = c(yearly_cases, yearly_deaths),
    names_glue = "{.value}_{year}"
  ) 


final_df
```


```{r}
final_df_inequality = final_df |> 
  mutate(
    inequality = health_80th_percentile_income / health_20th_percentile_income, 
    cases_percap_2020 = yearly_cases_2020 / population_2019,
    deaths_percap_2020 = yearly_deaths_2020 / population_2019,
    cases_percap_2023  = yearly_cases_2023 / population_2019,
    deaths_percap_2023 = yearly_deaths_2023 / population_2019,
    deaths_percap_2021 = yearly_deaths_2021 / population_2019,
    cases_percap_2021  = yearly_cases_2021 / population_2019,
    deaths_percap_2022 = yearly_deaths_2022 / population_2019,
    cases_percap_2022  = yearly_cases_2022 / population_2019,
    death_decline_pct_change = (yearly_deaths_2020 - yearly_deaths_2023) / yearly_deaths_2020 * 100,
    total_cases_percap = (yearly_cases_2020 + yearly_cases_2021 + yearly_cases_2022 + yearly_cases_2023)/ population_2019, 
    total_deaths_percap = (yearly_deaths_2020 + yearly_deaths_2021 + yearly_deaths_2022 + yearly_deaths_2023)/ population_2019,
  )

final_df_inequality
```


```{r}
ggplot(final_df_inequality,
       aes(x = inequality, y = total_cases_percap)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "darkblue") +
  labs(
    title = "Total COVID Cases Per Capita vs. Inequality Index",
    x = "Inequality Index (80th pct income / 20th pct income)",
    y = "Total Cases Per Capita"
  ) +
  theme_minimal()


ggplot(final_df_inequality,
       aes(x = inequality, y = total_deaths_percap)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "darkred") +
  labs(
    title = "Total COVID Deaths Per Capita vs. Inequality Index",
    x = "Inequality Index (80th pct income / 20th pct income)",
    y = "Total Deaths Per Capita"
  ) +
  theme_minimal()
```

```{r}
cases_long = final_df_inequality |> 
  pivot_longer(
    cols = starts_with("cases_percap_"),
    names_to = "year",
    names_prefix = "cases_percap_",
    values_to = "cases_percap"
  ) |> 
  mutate(year = as.factor(year))

deaths_long = final_df_inequality |> 
  pivot_longer(
    cols = starts_with("deaths_percap_"),
    names_to = "year",
    names_prefix = "deaths_percap_",
    values_to = "deaths_percap"
  ) |> 
  mutate(year = as.factor(year))


ggplot(cases_long, aes(x = inequality, y = cases_percap)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "darkred") +
  facet_wrap(~ year, ncol = 3) +
  labs(
    title = "Income Inequality vs COVID-19 Cases Per Capita (2020–2023)",
    x = "Income Inequality (80th / 20th percentile income)",
    y = "Cases Per Capita"
  ) +
  coord_cartesian(ylim = c(0, 0.50)) +
  theme_minimal()

ggplot(deaths_long, aes(x = inequality, y = deaths_percap)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "darkblue") +
  facet_wrap(~ year, ncol = 3) +
  labs(
    title = "Income Inequality vs COVID-19 Deaths Per Capita (2020–2023)",
    x = "Income Inequality (80th / 20th percentile income)",
    y = "Deaths Per Capita"
  ) +
  theme_minimal()
```

```{r}
ggplot(cases_long, aes(x = poverty_rate, y = cases_percap)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "darkred") +
  facet_wrap(~ year, ncol = 3) +
  labs(
    title = "Poverty Rate vs COVID-19 Cases Per Capita (2021–2023)",
    x = "Poverty Rate (%)",
    y = "Cases Per Capita"
  ) +
  coord_cartesian(ylim = c(0, 0.50)) +
  theme_minimal()

ggplot(deaths_long, aes(x = poverty_rate, y = deaths_percap)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "darkblue") +
  facet_wrap(~ year, ncol = 3) +
  labs(
    title = "Poverty Rate vs COVID-19 Deaths Per Capita (2021–2023)",
    x = "Poverty Rate (%)",
    y = "Deaths Per Capita"
  ) +
  theme_minimal()
```


```{r}
#Cases model with year interaction
lm_cases_all = lm(cases_percap ~ inequality * year, data = cases_long)
summary(lm_cases_all)

# Deaths model with year interaction
lm_deaths_all = lm(deaths_percap ~ inequality * year, data = deaths_long)
summary(lm_deaths_all)
```
The analysis examined the relationship between inequality and per capita COVID-19 cases and deaths across different years. For cases per capita, the overall effect of inequality was not statistically significant; however, the interaction between inequality and the year 2022 was strongly positive, indicating that inequality was associated with increased cases specifically in that year. Additionally, case rates differed significantly across years, with 2021 and 2022 showing higher rates compared to the baseline year, while 2023 showed a decline. In contrast, inequality had a significant positive association with deaths per capita overall, suggesting that higher inequality corresponded to higher death rates. This relationship varied by year: the effect of inequality on deaths was stronger in 2021 but weakened and even reversed in 2022 and 2023. These findings highlight that the impact of inequality on COVID-19 outcomes changed over time, emphasizing the importance of considering temporal dynamics in understanding health disparities during the pandemic.

## Race Analysis

```{r}
df_race_combined = final_df_inequality |> 
  mutate(
    race_hispanic = race_hispanic_male + race_hispanic_female,
    race_white = race_non_hispanic_white_alone_male + race_non_hispanic_white_alone_female,
    race_black = race_black_alone_male + race_black_alone_female,
    race_asian = race_asian_alone_male + race_asian_alone_female
  )
df_select = df_race_combined |> 
  select(county, state, starts_with("cases_percap_"), starts_with("deaths_percap_"),
         race_hispanic, race_white, race_black, race_asian)


df_long = df_select |> 
  pivot_longer(
    cols = starts_with(c("cases_percap_", "deaths_percap_")),
    names_to = c("measure", "year"),
    names_pattern = "(cases|deaths)_percap_(\\d{4})",
    values_to = "percap_value"
  )

df_long 
```

```{r}
library(dplyr)
library(heatmaply)

corr_matrix = cor(
  df_race_combined |> 
    select(race_white, race_black, race_asian, race_hispanic),
  df_race_combined |> 
    select(total_cases_percap, total_deaths_percap),
  use = "complete.obs"
)

heatmaply(
  corr_matrix,
  xlab = "Outcome Variables",
  ylab = "Race Variables",
  main = "Correlation between Race and COVID Outcomes",
  colors = colorRampPalette(c("red", "white", "blue"))(200),
  limits = c(-1, 1)
)

```


```{r}
df_final_pivot = df_long |> 
  pivot_longer(
    cols = c(race_hispanic, race_white, race_black, race_asian), 
    names_to = "race",
    values_to = "race_prop"
  ) |> 
  mutate(
    race = tools::toTitleCase(gsub("race_", "", race)),
    year = factor(year, levels = c("2020", "2021", "2022", "2023")),
    measure = factor(measure, levels = c("cases", "deaths"))
  )

plot_race_vs_outcome = function(data, measure_filter, year_filter) {
  ggplot(
    data |> filter(measure == measure_filter, year == year_filter),
    aes(x = race_prop, y = percap_value)
  ) +
    geom_point(alpha = 0.6) +
    geom_smooth(method = "lm", se = FALSE, color = "darkblue") +
    facet_wrap(~ race) +
    labs(
      title = paste0("COVID ", tools::toTitleCase(measure_filter), " Per Capita in ", year_filter,
                     " vs Population Proportion by Race"),
      x = "Population Proportion",
      y = paste0(tools::toTitleCase(measure_filter), " Per Capita")
    ) +
    theme_minimal()
}

```

```{r}
plot_race_vs_outcome(df_final_pivot, "deaths", "2020")
plot_race_vs_outcome(df_final_pivot, "deaths", "2021")
plot_race_vs_outcome(df_final_pivot, "deaths", "2022")
plot_race_vs_outcome(df_final_pivot, "deaths", "2023")
```

```{r}

results_deaths = df_final_pivot |> 
  filter(measure == "deaths") |>  
  group_by(race, year, measure) |> 
  do(broom::tidy(lm(percap_value ~ race_prop, data = .))) |> 
  filter(term == "race_prop") |> 
  ungroup()

sig_results_deaths = results_deaths |> 
  filter(p.value < 0.05)

cleaned_sig_results_deaths = sig_results_deaths |> 
  select(race, year, p.value)


cleaned_sig_results_deaths

```

```{r}

all_race_years = expand.grid(
  race = unique(cleaned_sig_results_deaths$race),
  year = unique(cleaned_sig_results_deaths$year)
)

signif_df = all_race_years |> 
  mutate(significant = FALSE) |> 
  left_join(cleaned_sig_results_deaths |>  mutate(significant = TRUE) |>  select(race, year, significant),
            by = c("race", "year")) |> 
  mutate(significant = ifelse(is.na(significant.y), significant.x, significant.y)) |> 
  select(race, year, significant)

ggplot(signif_df, aes(x = year, y = race, fill = significant)) +
  geom_tile(color = "white") +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "gray90"),
                    labels = c(" NOT Significant", "Significant")) +
  labs(
    title = "Significant Associations of Race by Year (Deaths)",
    x = "Year",
    y = "Race",
    fill = "Significance"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The analysis reveals that the proportion of the Asian population in counties was consistently and strongly associated with COVID-19 death rates across all years from 2020 to 2023, indicating persistent disparities affecting this group throughout the pandemic. In contrast, the Black population showed a significant association with COVID-19 deaths primarily during the early years of 2020 and 2021, with this relationship weakening in subsequent years. The Hispanic and White populations exhibited significant associations in 2020, 2022, and 2023, but not in 2021, suggesting that the impact of these populations on COVID-19 mortality fluctuated over time. These patterns suggest that the relationship between race and COVID-19 mortality was dynamic, reflecting changing social, economic, and public health factors over the course of the pandemic.
