---
title: "Interactive 3D Plot"
output: html_document
---

```{r echo= FALSE}
knitr::opts_chunk$set(
echo = FALSE,
message = FALSE,
warning = FALSE
)
```


```{r}
library(tidyverse)
library(plotly)
library(viridisLite)
```

```{r}
covid_shootings_2020 <- readRDS("covid_shootings_2020.rds")
```

```{r three_d_scatter, fig.width = 10, fig.height = 7}


covid_plot_data= 
  covid_shootings_2020 |>
  select(
    violent_crime_rate,
    homicides_per_100k,
    covid_deaths_per_100k_2020) |>
  drop_na()

fig= 
  plot_ly(
  data = covid_plot_data,
  x = ~violent_crime_rate,
  y = ~homicides_per_100k,
  z = ~covid_deaths_per_100k_2020,
  type = "scatter3d",
  mode = "markers",
  marker = list(
    size = 3,
    color = ~covid_deaths_per_100k_2020,
    colorscale = "Viridis",
    showscale = TRUE,
    colorbar = list(
      title = "COVID deaths per 100k (2020)",
      titleside = "right")),
  hovertemplate = paste(
    "Violent crime rate: %{x:.1f}<br>",
    "Homicides per 100k: %{y:.2f}<br>",
    "COVID deaths per 100k: %{z:.0f}<extra></extra>"))

fig= 
  fig |>
  layout(
    title = list(
      text = "3D Relationship Between Community Violence and COVID-19 Mortality",
      y = 0.92,          
      yanchor = "top",
      font = list(size = 20)),
    margin = list(l = 90, r = 90, b = 90, t = 40),   
    scene = list(
      xaxis = list(title = "Violent crime rate"),
      yaxis = list(title = "Homicides per 100k"),
      zaxis = list(title = "COVID-19 deaths per 100k (2020)"),
      aspectmode = "cube",
      camera = list(
        eye = list(x = 1.4, y = 1.4, z = 1.1))
    )
  )

fig
```

This interactive 3D scatterplot visualizes how county-level violent crime rates, homicide rates, and COVID-19 deaths relate across the United States. Each point represents a county, and the vertical axis and color scale show the intensity of COVID-19 mortality in 2020. This view allows us to observe how violence-related factors correspond with pandemic outcomes when considered simultaneously. The patterns that emerge in the visualization are summarized below.

```{r key_patterns, echo = FALSE}
library(tidyverse)
library(knitr)

patterns_df <- tibble(
  `Key pattern` = c(
    "Higher homicide rates are linked with higher COVID-19 mortality.",
    "Violent crime rate also increases with COVID-19 deaths, but more gradually.",
    "Counties low in both homicide and violent crime tend to have lower COVID-19 mortality.",
    "Looking at all three variables together helps show how violence levels relate to COVID-19 outcomes."
  )
)

kable(
  patterns_df,
  caption = "Key patterns observed in the 3D violence–COVID plot"
)

```

